Accurate Simulation of Flow through Dipping Aquifers with MODFLOW 6 Using Enhanced Cell Connectivity

### 2024-11-22 latex/article_clean_FINAL.tex
make figures more accessible
* change colors and symbols used in plots to make them more accessible for colorblind readers
* update manuscript accordingly

\begin{figure}
	\begin{center}
	\includegraphics[scale=0.6]{../figures/schem_conn_area_flux.png}
	\caption{Schematics showing (a) cell connectivity and (b) cell-cell interface fluxes in a two-grid-layer model of a dipping hydrogeologic layer. Hatching denotes impermeable boundaries along the top and bottom of the hydrogeologic layer. In (a), black circles represent cell centers. Blue lines represent hydraulic connections between cells in a layered-connectivity grid. Red lines represent additional connections that must be made to achieve full connectivity. In (b), gray arrows represent the uniform groundwater flux the model is attempting to simulate. Blue arrows represent components of the uniform groundwater flux normal to the horizontal and vertical cell-cell interfaces in the layered-connectivity grid. Red arrows represent components of the uniform groundwater flux normal to the additional vertical cell-cell interfaces introduced to represent the full connectivity.}
	\label{fig:schem-conn-area-flux}
	\end{center}
\end{figure}

Figure \ref{fig:schem-conn-area-flux}a illustrates what we refer to as "layered" and "full" cell connectivity. The focus is on a group of four cells in a two-grid-layer, vertically offset MODFLOW 6 model of fully saturated flow through a permeable hydrogeologic layer (aquifer) with impermeable top and bottom boundaries. If the grid is represented using a layered grid type (DIS or DISV), the grid has what we call ``layered connectivity": a cell is hydraulically connected to each laterally adjacent cell in the same grid layer (regardless of whether or not the cell faces actually overlap), and with each overlying and underlying cell, as indicated by the blue lines that connect cell centers in Figure \ref{fig:schem-conn-area-flux}a. In layered connectivity, two cells that have overlapping vertical faces but are in different grid layers (e.g., cells 1A and 2B) notably do not have a direct hydraulic connection. The addition of such connections between cells in different grid layers, as indicated by the red lines that connect cell centers in Figure \ref{fig:schem-conn-area-flux}a, results in a grid with what we call ``full connectivity.''  Implementation of full connectivity in the current version of MODFLOW 6 \citep{modflow650software} requires that the vertically offset grid be represented using the fully unstructured (DISU) grid type.

The need for full cell connectivity to accurately simulate flow along the dipping hydrogeologic layer is illustrated in Figure \ref{fig:schem-conn-area-flux}b. Components of the groundwater flux (specific discharge) vector are displayed on each cell-cell interface and are representative of steady, uniform flow along the hydrogeologic layer. Gray vectors represent the uniform flux oriented along the hydrogeologic layer, which the model is attempting to simulate. Blue vectors represent the flux components normal to cell-cell interfaces that correspond to blue connections in Figure \ref{fig:schem-conn-area-flux}a, which are common to both the layered-connectivity and full-connectivity grids. Red vectors represent the flux components normal to cell-cell interfaces that correspond to red connections in Figure \ref{fig:schem-conn-area-flux}a, which are lacking in the layered-connectivity grid.

The blue vectors in Figure \ref{fig:schem-conn-area-flux}b show that to simulate steady, uniform flow along the dipping hydrogeologic layer, cells must exchange water not only ``horizontally'' with neighboring cells in the same grid layer, but also vertically with cells in the adjacent grid layer. Specifically, there must be flow from cells in the lower grid layer to cells in the upper grid layer, which corresponds to the vertical component of the groundwater flux. Note, however, that the blue vectors alone are incompatible with steady, uniform flow along the hydrogeologic layer because the exchange of water between the two grid layers is in one direction only: upward. Such a steady-state flow configuration would imply continuous depletion of flow within the lower grid layer and accumulation of flow within the upper grid layer as one moves along the hydrogeologic layer in the direction of flow. Thus, the steady, uniform flow represented by the gray vectors cannot be simulated accurately without some mechanism for returning flow from the upper grid layer to the lower grid layer. The full-connectivity grid provides such a mechanism by offering additional connections between cells in adjacent grid layers. Vertical flows from the lower grid layer to the upper grid layer, represented by vertical blue vectors in Figure \ref{fig:schem-conn-area-flux}b, can be returned to the lower grid layer by nominally ``horizontal'' flows, represented by red vectors in Figure \ref{fig:schem-conn-area-flux}b. Although this concept has been illustrated using a two-grid-layer model for simplicity, the same reasoning applies given any number of grid layers.

@@

\begin{figure}[p!]
	\begin{center}
	\includegraphics[scale=0.75]{../figures/fig2_paper.png}
	\caption{Numerical results for the test problem using base-case settings for layered-connectivity and full-connectivity grids, with the standard conductance-based formulation as well as the XT3D formulation. The left panel of each scenario shows the calculated specific discharge at each cell center (arrows). The middle panel shows the face flows (arrows). The right panel shows head contours for the analytical solution (black) as well as the numerical solution (blue). Flux magnitudes, flux directions (angles), and head errors are reported to two decimal places. Percent errors in flow are reported to one decimal place.}
	\label{fig:fig2}
	\end{center}
\end{figure}

@@

\begin{figure}
	\begin{center}
	\includegraphics[scale=0.9]{../figures/fig3paper.png}
	\caption{Graphs indicating how the number of grid layers within the aquifer affects the flux magnitude in the center cell (top), the flux direction in center cell (middle) and the volumetric flow through the aquifer (bottom). Errors are minimized by using full connectivity (right column), the XT3D flow formulation (orange lines) and at least two grids layers in the aquifer.}
	\label{fig:fig3}
	\end{center}
\end{figure}

@@

The base case uses an aquifer dip angle of 30$^{\circ}$ and a conductivity contrast between domain and aquifer of $10^{-6}$:1. In this subsection, we investigate the behavior of the flow solution for aquifer dip angles ranging from 0 to 80$^{\circ}$. We also examine the behavior for a range of conductivity contrasts by setting the domain conductivity to 1/2, 1/5, 1/10 and 1/100 that of the aquifer. Simulated flux magnitude and direction at the center cell for all combinations are plotted against the analytical solution in Figure \ref{fig:fig4}.

@@

\caption{Flux magnitude and direction at the cell in the center of the aquifer as a function of aquifer dip angle for various ratios of domain to aquifer conductivity (blue, orange, green, red lines). The analytical solution is shown as a black dashed line.}
\label{fig:fig4}
\end{figure}

@@

Layered-connectivity grids produce results that deviate substantially from the analytical solution both with and without the XT3D formulation (Figure \ref{fig:fig5}a), whereas full-connectivity grids with XT3D reproduce the analytical solution (Figure \ref{fig:fig5}b, orange lines). For example, at an anisotropy ratio of 0.01, use of layered connnectivity produces flow errors of approximately -96\% for both the standard and XT3D flow formulations. Use of full connectivity produces errors of approximately 30\% and -3\% for the standard and XT3D flow formulations, respectively. These results suggest that use of full grid connectivity and the XT3D flow formulation in MODFLOW 6 simulations can be particularly important in the presence of anisotropy.

\begin{figure}
	\begin{center}
	\includegraphics[scale=0.9]{../figures/fig5paper.png}
	\caption{Graphs indicating how the anisotropy ratio within the aquifer and domain affects the flux magnitude in the center cell (top), the flux direction in center cell (middle) and the volumetric flow through the aquifer (bottom). Errors are minimized by using full connectivity (right column) and the XT3D flow formulation (orange lines).}
	\label{fig:fig5}
	\end{center}
\end{figure}

>>>>

\begin{figure}
	\begin{center}
	\includegraphics[scale=0.5]{../figures/fig1_paper.png}
	\caption{Schematics showing (a) cell connectivity and (b) cell-cell interface fluxes in a two-grid-layer model of a dipping hydrogeologic layer. Hatching denotes impermeable boundaries along the top and bottom of the hydrogeologic layer. In (a), black circles represent cell centers. Solid blue lines represent hydraulic connections between cells in a layered-connectivity grid. Dashed red lines represent additional connections that must be made to achieve full connectivity. In (b), solid gray arrows represent the uniform groundwater flux the model is attempting to simulate. Solid blue arrows represent components of the uniform groundwater flux normal to the horizontal and vertical cell-cell interfaces in the layered-connectivity grid. Dashed red arrows represent components of the uniform groundwater flux normal to the additional vertical cell-cell interfaces introduced to represent the full connectivity.}
	\label{fig:schem-conn-area-flux}
	\end{center}
\end{figure}

Figure \ref{fig:schem-conn-area-flux}a illustrates what we refer to as "layered" and "full" cell connectivity. The focus is on a group of four cells in a two-grid-layer, vertically offset MODFLOW 6 model of fully saturated flow through a permeable hydrogeologic layer (aquifer) with impermeable top and bottom boundaries. If the grid is represented using a layered grid type (DIS or DISV), the grid has what we call ``layered connectivity": a cell is hydraulically connected to each laterally adjacent cell in the same grid layer (regardless of whether or not the cell faces actually overlap), and with each overlying and underlying cell, as indicated by the solid blue lines that connect cell centers in Figure \ref{fig:schem-conn-area-flux}a. In layered connectivity, two cells that have overlapping vertical faces but are in different grid layers (e.g., cells 1A and 2B) notably do not have a direct hydraulic connection. The addition of such connections between cells in different grid layers, as indicated by the dashed red lines that connect cell centers in Figure \ref{fig:schem-conn-area-flux}a, results in a grid with what we call ``full connectivity.''  Implementation of full connectivity in the current version of MODFLOW 6 \citep{modflow650software} requires that the vertically offset grid be represented using the fully unstructured (DISU) grid type.

The need for full cell connectivity to accurately simulate flow along the dipping hydrogeologic layer is illustrated in Figure \ref{fig:schem-conn-area-flux}b. Components of the groundwater flux (specific discharge) vector are displayed on each cell-cell interface and are representative of steady, uniform flow along the hydrogeologic layer. Gray vectors represent the uniform flux oriented along the hydrogeologic layer, which the model is attempting to simulate. Solid blue vectors represent the flux components normal to cell-cell interfaces that correspond to solid blue connections in Figure \ref{fig:schem-conn-area-flux}a, which are common to both the layered-connectivity and full-connectivity grids. Dashed red vectors represent the flux components normal to cell-cell interfaces that correspond to dashed red connections in Figure \ref{fig:schem-conn-area-flux}a, which are lacking in the layered-connectivity grid.

The solid blue vectors in Figure \ref{fig:schem-conn-area-flux}b show that to simulate steady, uniform flow along the dipping hydrogeologic layer, cells must exchange water not only ``horizontally'' with neighboring cells in the same grid layer, but also vertically with cells in the adjacent grid layer. Specifically, there must be flow from cells in the lower grid layer to cells in the upper grid layer, which corresponds to the vertical component of the groundwater flux. Note, however, that the solid blue vectors alone are incompatible with steady, uniform flow along the hydrogeologic layer because the exchange of water between the two grid layers is in one direction only: upward. Such a steady-state flow configuration would imply continuous depletion of flow within the lower grid layer and accumulation of flow within the upper grid layer as one moves along the hydrogeologic layer in the direction of flow. Thus, the steady, uniform flow represented by the gray vectors cannot be simulated accurately without some mechanism for returning flow from the upper grid layer to the lower grid layer. The full-connectivity grid provides such a mechanism by offering additional connections between cells in adjacent grid layers. Vertical flows from the lower grid layer to the upper grid layer, represented by vertical solid blue vectors in Figure \ref{fig:schem-conn-area-flux}b, can be returned to the lower grid layer by nominally ``horizontal'' flows, represented by dashed red vectors in Figure \ref{fig:schem-conn-area-flux}b. Although this concept has been illustrated using a two-grid-layer model for simplicity, the same reasoning applies given any number of grid layers.

@@

\begin{figure}[p!]
	\begin{center}
	\includegraphics[scale=0.75]{../figures/fig2_paper.png}
	\caption{Numerical results for the test problem using base-case settings for layered-connectivity and full-connectivity grids, with the standard conductance-based formulation as well as the XT3D formulation. The left panel of each scenario shows the calculated specific discharge at each cell center (arrows). The middle panel shows the face flows (arrows). The right panel shows head contours for the analytical solution (dashed black) as well as the numerical solution (solid blue). Flux magnitudes, flux directions (angles), and head errors are reported to two decimal places. Percent errors in flow are reported to one decimal place.}
	\label{fig:fig2}
	\end{center}
\end{figure}

@@

\begin{figure}
	\begin{center}
	\includegraphics[scale=0.9]{../figures/fig3_paper.png}
	\caption{Graphs indicating how the number of grid layers within the aquifer affects the flux magnitude in the center cell (top), the flux direction in center cell (middle) and the volumetric flow through the aquifer (bottom). Errors are minimized by using full connectivity (right column), the XT3D flow formulation (orange lines with solid circles) and at least two grids layers in the aquifer.}
	\label{fig:fig3}
	\end{center}
\end{figure}

@@

The base case uses an aquifer dip angle of 30$^{\circ}$ and a conductivity contrast between domain and aquifer of $10^{-6}$:1. In this subsection, we investigate the behavior of the flow solution for aquifer dip angles ranging from 0 to 80$^{\circ}$. We also examine the behavior for a range of conductivity contrasts by setting the domain conductivity equal to and to 1/2, 1/5, 1/10 and 1/100 that of the aquifer. Simulated flux magnitude and direction at the center cell for all combinations are plotted against the analytical solution in Figure \ref{fig:fig4}.

@@

\caption{Flux magnitude and direction at the cell in the center of the aquifer as a function of aquifer dip angle for various ratios of domain to aquifer conductivity (solid and dash-dotted lines of varying color). The analytical solution is shown as a dashed black line.}
\label{fig:fig4}
\end{figure}

@@

Layered-connectivity grids produce results that deviate substantially from the analytical solution both with and without the XT3D formulation (Figure \ref{fig:fig5}a), whereas full-connectivity grids with XT3D reproduce the analytical solution (Figure \ref{fig:fig5}b). For example, at an anisotropy ratio of 0.01, use of layered connnectivity produces flow errors of approximately -96\% for both the standard and XT3D flow formulations. Use of full connectivity produces errors of approximately 30\% and -3\% for the standard and XT3D flow formulations, respectively. These results suggest that use of full grid connectivity and the XT3D flow formulation in MODFLOW 6 simulations can be particularly important in the presence of anisotropy.

\begin{figure}
	\begin{center}
	\includegraphics[scale=0.9]{../figures/fig5_paper.png}
	\caption{Graphs indicating how the anisotropy ratio within the aquifer and domain affects the flux magnitude in the center cell (top), the flux direction in center cell (middle) and the volumetric flow through the aquifer (bottom). Errors are minimized by using full connectivity (right column) and the XT3D flow formulation (orange lines with solid circles).}
	\label{fig:fig5}
	\end{center}
\end{figure}
